---
title: "Regional Analysis of Preterm Birth Characteristics""
subtitle: "BMIN503/EPID600 Final Project"
author: "Rose Albert"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

# Interrogating sex as a biological variable in county-level preterm birth inequity

December 13, 2024

By Rose Albert

## Overview {#sec-overview}

Preterm birth affects approximately 10% of all births, and there is a well-established male vulnerability and female resilience during the neonatal period. The goal of this project is to interrogate sex as a biological variable at the county-level using CDC Wonder Natality Data and identify sex-differences in preterm birth rates and NICU admission. I have spoken with Dr. Aimin Chen and Dr. Heather Burris about biological and structural determinants contributing to preterm birth inequity, as well as potential areas of clinical, public health, and policy intervention.

The data for this project can be found in my [final project GitHub repository](https://github.com/rosemalbert/BMIN503_Final_Project/tree/master).

## Introduction {#sec-introduction}

Preterm birth can predispose infants to immediate risks in the neonatal period as well as long-term adverse sequelae. For example, preterm infants are born with underdeveloped lungs and often require therapeutic interventions such as surfactant administration, antibiotics, steroids, mechanical ventilation, and supplemental oxygen for survival. However, these therapeutics can also pose postnatal insults and contribute to lung injury and disease such as bronchopulmonary dysplasia.

Interrogation of sex as a biological variable has received increasing attention as a critical area for understanding basic pathophysiology and differential response to therapeutics. Male sex is an independent risk factor for numerous pediatric diseases and diseases of prematurity. This project will investigate male susceptibility to preterm birth and NICU admission at the county-level. An understanding of preterm birth characteristics can inform local public health interventions.

## Methods {#sec-methods}

[*Data curation and cleaning*]{.underline}

*County-level Birth Characteristics Data Source:* I am using the [CDC Wonder Natality Data](https://wonder.cdc.gov/wonder/help/Natality-expanded.html#) (2023) grouped by "County of Residence," "Sex of Infant," "NICU Admission," and "OE Gestational Age Recode 10," with the measures "Births" and "Percent of Total Births."

*Mapping Data Source:* I am drawing from Assignment 5 to use geospatial data, tidy census, and leaflet to generate interactive maps.

```{r}
#load necessary libraries
library(tidyverse)
library(sf)
library(tigris)
library(tidycensus)
library(leaflet)
options(tigris_use_cache = TRUE)
options(progress_enabled = FALSE)

#Load county-level birth data downloaded from CDC Wonder
birth_data <- read.delim("https://raw.githubusercontent.com/rosemalbert/BMIN503_Final_Project/refs/heads/master/Natality%2C%202016-2023%20expanded_sex.txt")

county_data <- read.delim("https://raw.githubusercontent.com/rosemalbert/BMIN503_Final_Project/refs/heads/master/Natality%2C%202016-2023%20expanded%20_county.txt")

#view column names 
colnames(birth_data)
colnames(county_data)

#merge birth_data and county_data

#calculate birth rate for each characteristic and add as a column 

#load census shapefile data and merge by county FIPs 
```

[*Analysis plan*]{.underline}

*Exploratory analysis of data:* Generate county-level choropleth maps of birth rates (births/total population for each county). Generate county-level choropleth maps of preterm birth rates (\<37 gestational age births/total population). Generate sex-stratified maps of preterm birth rates. Generate sex-stratified choropleth maps of NICU admissions by county. Generate bar graphs showing NICU admission rates by sex and sex distribution of gestational age groups.

*Data analysis:* Conduct 2-way ANOVA to determine if sex as a biological variable is a significant predictor of NICU admission. Identify counties with highest and lowest preterm birth rates, and the highest and lowest NICU admission rates.

## Results {#sec-results}

```{r}
library(dplyr)


# For a tab-delimited text file (common format for CDC data):
data <- read.delim("https://raw.githubusercontent.com/rosemalbert/BMIN503_Final_Project/refs/heads/master/Natality%2C%202016-2023%20expanded_sex.txt")
glimpse(data)

```

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked here.
